R Notebook

### 2017-04-05 doc/RMD/MLG-distribution.Rmd
update analysis to weight by psex and include notes

I realized that it's possible to use an MLG table with matrix multiplication to
get an adjency matrix.

```{r}
make_graph_list <- function(dat){ # dat is a genclone/snpclone object
  datmlg  <- mlg.table(dat, plot = FALSE) > 0 # presence/absence of MLG
  crosses <- mlg.crosspop(dat, quiet = TRUE, df = TRUE) %>% tbl_df()
  adjmat  <- datmlg %*% t(datmlg) 
  cols    <- sort(colnames(adjmat))
  adjmat  <- adjmat[cols, cols]
  # Creating Graph
  g           <- graph_from_adjacency_matrix(adjmat, mode = "undirected", diag = FALSE)
  V(g)$size   <- diag(adjmat)
  g           <- delete_vertices(g, degree(g) == 0)
  shared_mlg  <- (crosses %>% group_by(Population) %>% summarize(n = n()))$n
  V(g)$weight <- 1 - shared_mlg/V(g)$size
  el          <- as_adj_edge_list(g)
  el          <- el[lengths(el) > 0]
  popgraphs <- setNames(vector(mode = "list", length = length(el) + 1), c(names(el), "total"))
  for (v in names(el)){
    idx  <- el[[v]]
    mlgs <- crosses %>%           # How to get all MLGs from a single population:
      filter(Population == v) %>%         # Grab only the population e and then
      select(MLG) %>%                     # remove everything but the MLGs to do an
      inner_join(crosses, by = "MLG") %>% # inner join of the original list and then
      filter(Population != v) %>%         # remove the query population to give
      arrange(Population)                 # the neigboring populations in order.
    MLGS <- as.character(mlgs$MLG)
    E(g)[idx]$label  <- substr(MLGS, 5, nchar(MLGS))
    E(g)[idx]$weight <- as.integer(table(MLGS)[MLGS]) # weight == n populations visited
    popgraphs[[v]]   <- subgraph.edges(g, eids = idx)
  }
  popgraphs[["total"]] <- g
  popgraphs
}

plot_mlg_graph <- function(g, glayout = NULL){
  if (is.null(glayout)){
    glayout <- layout_nicely(g)
  } else {
    glayout <- glayout[V(g)$name, ]
  }
  shared_mlg  <- (1 - V(g)$weight) * V(g)$size
  g           <- add_vertices(g, length(V(g)), size = V(g)$size - shared_mlg, color = "grey90")
  g2          <- g
  V(g2)$label <- ifelse(!is.na(V(g2)$name), sprintf("%s [(%d/%d)]", V(g2)$name, shared_mlg, V(g2)$size), NA)
  glay        <- create_layout(g2, "manual", node.positions = as.data.frame(rbind(glayout, glayout)))
  x_nudge     <- ifelse(abs(glay$x) == 1, -glay$x/10, glay$x/10)
  ggraph(glay) +
    geom_edge_fan(aes(alpha = weight + 1), width = 0.75) +
    geom_node_circle(aes(r = drop(scale(size, center = FALSE))/10, fill = size)) +
    geom_node_label(aes(label = label), repel = TRUE, parse = TRUE, label.size = 0.75, nudge_x = x_nudge) +
    viridis::scale_fill_viridis(option = "C") +
    scale_edge_alpha_continuous(range = c(0.25, 1), breaks = c(2:5)) +
    coord_fixed() +
    theme_void() +
    theme(text = element_text(size = 14)) +
    labs(list(
      title = "Shared haplotypes across regions",
      fill = "Number of\nGenotypes",
      edge_alpha = "Populations\nper haploytpe",
      caption = "Outer circle: Number of haplotypes in the region\nInner Circle: Number of private haplotypes in the region"
    ))
}

plot_mlg_subgraph <- function(graphlist){
  for (i in names(graphlist)){
    pg   <- graphlist[[i]]
    labs <- ifelse(E(pg)$weight > 1, E(pg)$label, NA)
    labs <- ifelse(duplicated(labs), NA, labs)
    plot(pg, 
         main = i, 
         layout = layout_as_star(pg, center = i), 
         edge.width = E(pg)$weight,
         edge.label = labs)
  }
}

@@

Now that we have the adjacency matrix, we can use it to construct our graph:

```{r, fig.width = 7, fig.height = 7}
graph16loc <- make_graph_list(dat)
graph11loc <- make_graph_list(dat[loc = keeploci, mlg.reset = TRUE])

par(mfrow = c(3, 4))
plot_mlg_subgraph(graph16loc[-length(graph16loc)])

par(mfrow = c(3, 4))
plot_mlg_subgraph(graph11loc[-length(graph11loc)])

par(mfrow = c(1, 1))
plot_mlg_graph(graph16loc$total, good_layout)
plot_mlg_graph(graph11loc$total, good_layout)
```

```{r, eval = FALSE}
null_vertices <- length(V(g)) %>% seq(from = ./2 + 1, to = ., by = 1)
sg <- g %>% 
  delete_vertices(null_vertices) %>%
  cluster_spinglass()
modularity(sg)
communities(sg)
```

>>>>

I realized that it's possible to use an MLG table with matrix multiplication to
get an adjency matrix. Here I'm making a function to create a list of graphs.

The vertices, representing populations, will have the following attributes:

 - Size: equal to the number of MLGs within the population
 - Weight: fraction of private MLGs

The edges, representing multilocus genotypes are a bit trickier to handle since
multiple edges can represent a single MLG passing through populations. They have
the following attributes:

 - Label: the MLG this edge belongs to. Note that multiple edges can share the
          same label.
 - Width: the number of populations this MLG passes through
 - Weight: 1 - (probability of a second encounter by chance); 1 - psex

The reason why I'm weighting these nodes as 1 - psex is for subsequent community
analysis. The more likely you are to encounter a genotype by chance, the less
influence that genotype should have on the connectivity.

```{r}
rerange <- function(x){
  minx <- min(x, na.rm = TRUE)
  maxx <- max(x, na.rm = TRUE)
  if (minx < 0)
    x <- x + abs(minx)
    maxx <- maxx + abs(minx)
  if (maxx > 1)
    x <- x/maxx
  return(x)
}
make_graph_list <- function(dat){ # dat is a genclone/snpclone object
  # w       <- pgen(dat, by_pop = FALSE) %>% rowSums() %>% exp() %>% setNames(paste0("MLG.", mll(dat)))
  # w       <- w[unique(names(w))]
  w       <- psex(dat, by_pop = FALSE, method = "multiple") %>% split(mll(dat))
  names(w)<- paste0("MLG.", names(w))
  datmlg  <- mlg.table(dat, plot = FALSE) > 0 # presence/absence of MLG
  crosses <- mlg.crosspop(dat, quiet = TRUE, df = TRUE) %>% tbl_df()
  adjmat  <- datmlg %*% t(datmlg) 
  cols    <- sort(colnames(adjmat))
  adjmat  <- adjmat[cols, cols]
  # Creating Graph
  g           <- graph_from_adjacency_matrix(adjmat, mode = "undirected", diag = FALSE)
  V(g)$size   <- diag(adjmat)
  g           <- delete_vertices(g, degree(g) == 0)
  shared_mlg  <- (crosses %>% group_by(Population) %>% summarize(n = n()))$n
  V(g)$weight <- 1 - shared_mlg/V(g)$size # fraction of private MLGs
  el          <- as_adj_edge_list(g)
  el          <- el[lengths(el) > 0]
  popgraphs <- setNames(vector(mode = "list", length = length(el) + 1), c(names(el), "total"))
  for (v in names(el)){
    idx  <- el[[v]]
    mlgs <- crosses %>%           # How to get all MLGs from a single population:
      filter(Population == v) %>%         # Grab only the population e and then
      select(MLG) %>%                     # remove everything but the MLGs to do an
      inner_join(crosses, by = "MLG") %>% # inner join of the original list and then
      filter(Population != v) %>%         # remove the query population to give
      arrange(Population)                 # the neigboring populations in order.
    MLGS <- as.character(mlgs$MLG)
    E(g)[idx]$label  <- substr(MLGS, 5, nchar(MLGS))
    E(g)[idx]$width  <- as.integer(table(MLGS)[MLGS]) # size == n populations visited
    E(g)[idx]$weight <- 1 - map_dbl(w[MLGS], 2)       # weight == 1 - psex 2nd encounter
    popgraphs[[v]]   <- subgraph.edges(g, eids = idx)
  }
  popgraphs[["total"]] <- g
  popgraphs
}

plot_mlg_graph <- function(g, glayout = NULL){
  if (is.null(glayout)){
    glayout <- layout_nicely(g)
  } else {
    glayout <- glayout[V(g)$name, ]
  }
  shared_mlg  <- (1 - V(g)$weight) * V(g)$size
  g           <- add_vertices(g, length(V(g)), size = V(g)$size - shared_mlg, color = "grey90")
  g2          <- g
  V(g2)$label <- ifelse(!is.na(V(g2)$name), sprintf("%s [(%d/%d)]", V(g2)$name, shared_mlg, V(g2)$size), NA)
  glay        <- create_layout(g2, "manual", node.positions = as.data.frame(rbind(glayout, glayout)))
  x_nudge     <- ifelse(abs(glay$x) == 1, -glay$x/10, glay$x/10)
  breaks      <- quantile(1 - E(g)$weight)
  breaks      <- setNames(breaks, format(breaks, digits = 2))
  ggraph(glay) +
    geom_edge_fan(aes(alpha = 1 - weight, width = width + 1)) +
    geom_node_circle(aes(r = drop(scale(size, center = FALSE))/10, fill = size)) +
    geom_node_label(aes(label = label), repel = TRUE, parse = TRUE, label.size = 0.75, nudge_x = x_nudge) +
    viridis::scale_fill_viridis(option = "C") +
    scale_edge_alpha_continuous(range = c(1, 0.25), breaks = breaks) +
    scale_edge_width_continuous(range = c(0.25, 1.25), breaks = c(2:5)) +
    coord_fixed() +
    theme_void() +
    theme(text = element_text(size = 14)) +
    labs(list(
      title = "Shared haplotypes across regions",
      fill = "Number of\nGenotypes",
      edge_alpha = "Probability of\nsecond encounter",
      edge_width = "Populations\nPer haplotype",
      caption = "Outer circle: Number of haplotypes in the region\nInner Circle: Number of private haplotypes in the region"
    ))
}

plot_mlg_subgraph <- function(graphlist){
  for (i in names(graphlist)){
    pg   <- graphlist[[i]]
    labs <- ifelse(E(pg)$width > 1, E(pg)$label, NA)
    labs <- ifelse(duplicated(labs), NA, labs)
    plot(pg, 
         main = i, 
         layout = layout_as_star(pg, center = i), 
         edge.width = E(pg)$width,
         edge.color = grey(rerange(1 - E(pg)$weight)), 
         edge.label = labs)
  }
}

@@

## Graphs

With the fuctions above, we can create and plot the graphs.

```{r}
# Creating the graphs
graph16loc <- make_graph_list(dat)
graph11loc <- make_graph_list(dat[loc = keeploci, mlg.reset = TRUE])
```

### Individual population subgraphs

Each subgraph shows all of the connections for a single population.

```{r, fig.width = 10, fig.height = 10}
# Plotting the subgraphs
par(mfrow = c(3, 4))
plot_mlg_subgraph(graph16loc[-length(graph16loc)])

par(mfrow = c(3, 4))
plot_mlg_subgraph(graph11loc[-length(graph11loc)])
```

### Cross-regional graphs

#### With 16 loci

First, I'm going to show the results of a community analysis. I'm using the 
igraph function `cluster_optimal()` to cluster the nodes.

```{r, fig.width = 7, fig.height = 7}
par(mfrow = c(1, 1))
(g16o <- cluster_optimal(graph16loc$total))
plot_mlg_graph(graph16loc$total, good_layout) + labs(list(subtitle = "(16 loci)"))
```

What we see is that We are given three clusters showing a clustering of the
plains states, the west coast, and Australia, France, and Minnesota. The last
cluster is likely driven by the single genotype shared between these three
regions that has a low probability of a second encounter.

#### With 11 loci

```{r, fig.width = 7, fig.height = 7}
(g11o <- cluster_optimal(graph11loc$total))
plot_mlg_graph(graph11loc$total, good_layout) + labs(list(subtitle = "(11 loci)"))
```
